---
title: "rasterTools - Introduction"
author: "Steffen Ehrmann"
date: "`r Sys.Date()`"
bibliography: rasterTools.bib
link-citations: yes
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

You might find `rasterTools` useful when your work depends in one way or the other on earth observation data [@bush_connecting_2017]. This may be for simulating spatial patterns, building species distribution models or analysing and measuring spatial patterns/landscape features.

`rasterTools` does NOT intend to replace `raster` [@hijmans_raster_2017] but tries to present an additional workflow that is aimed at being more transparent and reproducible and easily accessible. Many of the typical steps in working with raster data are summarised into *operators* (functions that carry out a specific spatial operation) and are grouped according to the group of tasks. In `rasterTools` the four functions `obtain()`, `generate()`, `modify()` and `measure()` represent the major tasks of spatial operations. `rasterTools` has been conceptualised so that these core functions are modular. Each of these functions manages all the code-logic that is common for the task at hand, e.g. to obtain datasets. More specific code that manages, for instance, "loading the MODIS dataset" is oursourced to a seperate function (the `oMODIS()` operator). New operators to all the core functions and to `loadData()` can hence be devised easily.

Several operators are combined to *algorithms* to carry out (a sequence of) more complex spatial operations. Operators are scope specific, i.e. they are typically only called in their respective core function. An algorithm in `rasterTools` is given as list of lists.

    algo_name <- list(list(operator1),
                      list(operator2))
                      
Each operator represents a `do.call()` compliant definition (type `?do.call` into the console to learn about it).

    do.call(what, args, quote = FALSE, envir = parent.frame())

Which translates to

    algo_name <- list(operator1 = list(what, args),
                      operator2 = list(what, args))

Its first element `what = operator_name` would outline the name of the function utilized in this operator (such as `"oMODIS"` or `"oCLC"`). Its second element (`args`) can have any number of arguments required to put together the call (such as `product = "MOD17A3", layer = 2, period = c(2012, 2014)` or simply `period = 2006`).

    datasets <- list(list(operator = "oCLC", period = 2006),
                     list(operator = "oGFC", period = c(2005:2007)))

This modularity is one of the strong sides of `rasterTools`. It allows the user to write their own operator to be used in the respective core-function. An example would be a dataset that is not yet supported by `rasterTools`, to utilize a modification which is not yet included in this package or to develop new landscape metrics. All of this can easily and seamlessly be combined with the previously defined operators in the same, simple workflow.

## A new (spatial) class: `geom`
`rasterTools` comes with its own (spatial) S4 class. This class is radically simplified (even in comparison to `sf`) and thus consists only of the feature classes `point`, `line` and `polygon`. It does, however, come with many functions that create specific types of lines or polygons, for instance `geomHexagon()`, which creates exclusively hexagonal polygon geometries.

These geometries are deliberately per default not *spatial* in nature and do also work within an ordinary cartesian coordinate system. They can be transformed into a spatial object by assigning a *coordinate reference system*. We learn a lot about them by simply creating and plotting one:

```{r}
library(raster)
library(rasterTools)

coords <- data.frame(x = c(40, 70, 70, 50),
                     y = c(40, 40, 60, 70),
                     fid = 1)
window <- data.frame(x = c(0, 80),
                     y = c(0, 80))
(aGeom <- geomPolygon(anchor = coords, window = window, show = TRUE))
```

The main feature of this sort of "spatial" class is that it can be created relatively simply by providing a set of coordinates as socalled anchor values. It has a new slot, `window`, which defines the frame within which it would be plotted. A wide range of "getters" and "setters" (for example `getExtent()` or `setCRS()`) have been defined as methods for `geom` and other spatial classes, in the latter case they are mostly a simple wrapper of the original `raster`, `sp` or `sf` funciton. The foreign methods have been defined for the sake of internal consistency and simplification.

Additionally, several "tools" to modify the `geom` vertices and to transform the `geom` to and from other spatial classes have been programmed (for example `gRotate()` or `gToSp()`) and this will be extended in the future.


# Example workflow
In the following you will find an example for each of the core functions.

`rasterTools` may write large files to your harddisc and you should assign a specific directory on a storage medium that has sufficient space via `setPaths(root = "/path/to/the/spatial/files")`.

## Generate spatial patterns
*In the making*

## Obtain spatial datasets
An algorithm with which one would obtain information from various spatial data sets could be:

```{r}
myDatasets <- list(list(operator = "oGFC", period = c(2006)),
                   list(operator = "oMODIS", product = "mod17a3", period = 2006,
                        layer = 2))
```

Most of the different spatial datasets are available in a specific file format and may have other specific properties, which require consideration. For instance, MODIS data are stored as products with several layers within each file. Consequently the `product` and `layer` need to be specified. Each of the required arguments can be found in the respective documentation, for instance with `?oMODIS`.

Typically we only need a (vastly) smaller subset of an earth observation dataset and we would want to use a rectangular mask (of the class `geom`) to outline this area of interest:

```{r}
myMask <- rtGeoms$mask
```

The collection of datasets for the area of our interest can be obtained by the following code:

```{r eval=FALSE}
myData <- obtain(data = myDatasets, mask = myMask)
```

(output here not shown as it may download the datasets, if you do not have them yet)

## Modify raster objects
A typical modification algorithm determines patches of the foreground in a raster with continuous integer values:

```{r}
get_patches <- list(list(operator = "rBinarise", thresh = 30),
                    list(operator = "rPatches"))
```

When using this algorithm in `modify()` while `sequential = TRUE` the binarised raster will be used to derive patches. The output of `rBinarise` will be passed to `rPatches`, which determines patches from the thresholded rasters values.

However, if you want to determine patches, but also derive categories from the original raster, you have to give each operator in the algorithm a specific name. Operators which should be part of the same sub-algorithm have to have the same name:

```{r}
cc_cats <- list(get_patches = list(operator = "rBinarise", thresh = 30),
                get_patches = list(operator = "rPatches"),
                get_categories = list(operator = "rCategorise", n = 5))
```

In this case `sequential = TRUE` would be ignored globally, but applied locally for those functions that share the same name (also if it would be set to `FALSE` in the call).

The modifications would be carried out by:

```{r, fig.width=7, out.width='100%'}
myInput <- rtData$continuous
myPatches <- modify(input = myInput, by = get_patches, sequential = TRUE)
visualise(raster::stack(myInput, myPatches))
```


## Measure raster objects
Landscape metrics are another case of algorithms in `rasterTools`. We distinguish between _generic_ and _derived_ landscape metrics. To evaluate a generic metric, typically just the respective operator with its arguments needs to be given:

```{r}
myInput <- rtData$categorical
myMetrics <- list(a_c = list(operator = "mArea", scale = "class"),
                  a_l = list(operator = "mArea", scale = "landscape"))
measure(input = myInput, with = myMetrics)
```

The so defined *terms* are the basis for derived metrics. Such a metric is defined simply by its mathematical equation, for example the *Class proportional area* is the class areas divided by the landscape area:

```{r}
myMetric <- list(a_l = list(operator = "mArea", scale = "landscape"),
                 a_c = list(operator = "mArea", scale = "class"),
                 mCPA = "a_c / a_l * 100")
```

Several derived metrics can be defined in one algorithm, given the required terms are defined:

```{r}
myMetrics <- list(a_p = list(operator = "mArea", scale = "patch"),
                  a_c = list(operator = "mArea", scale = "class"),
                  a_l = list(operator = "mArea", scale = "landscape"),
                  mCPA = "a_c / a_l * 100",
                  mLPI = "max(a_p) / a_l * 100")
```

The generic metrics `mNumber()`, `mPerimeter()`, `mArea()`, `mAdjacency()` and `mValues()` are available. Equations for the derived metrics can be found in the [landscape metrics](articles/landscape_metrics.html) vignette (for example the *largest patch index* `mLPI` or the *percentage of like adjacencies* `mPLA` and with increasing version number more of them). It becomes apparent that also new, possibly improved landscapes metrics can be easily prototyped and tested in this framework.

Finally, the measurement of the *class proportional area* and the *largest patch index*, in a raster where patches have been determined, would be carried out by:

```{r}
measure(input = myInput, with = myMetrics)
```


# Acknowledgements
I am grateful for financial support from the PROFOUND Cost-action, which gave me the opportunity to work in a concentrated effort a large part of the functionality. This package has been developed in support of the [FunBo Project](https://www.researchgate.net/project/Do-small-mammals-think-big-The-multiscale-ecology-of-small-mammals-and-their-functional-role-for-Borrelia-burgdorferi-FunBo), which was made possible by the Grünewald-Zuberbier Scholarship handed out by the University of Freiburg.

Thanks are also due to Prof. Arne Pommerening who was a great source of inspiration for what `rasterTools` is now.

# References